Region-Network Hierarchical Sparsity Priors for\\ High-Dimensional Inference in Brain Imaging

### 2016-02-04 paper/icml2016/example_paper.tex
make a and b explicit

\begin{figure*}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.55\textwidth]{../figures/ROC_ALL.png}
\vspace{-0.4cm}
\caption{\textbf{Prediction performance across sparsity priors.}
Comparison of the performance of logistic regression estimators
with 6 different structured and unstructered sparsity penalties
in classifying neural activity from 18 psychological tasks.
The class-wise area under the curve (AUC, \textit{colors})
is obtained on an identical test set.
%
\textit{Hierarchical Tree Sparsity}: Structured
$\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors
exhibited the best out-of-sample performance.
\textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
imposed a minimum of relevant brain voxels without
assuming special structure.
\textit{Elastic-Net}: Unstructured logistic regression
with interpolation between $\ell_1$- and $\ell_2$-norm 
imposed an equilibrium between sparsity and model fit.
\textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
norm (with additional $\ell_1$ term)
imposed region compartments, but naive to network structure.
\textit{Trace-norm}: Structured trace-norm penalization
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\end{centering}
\label{fig_sparsities}
\end{figure*}

% \paragraph{Benchmarking hierarchical tree sparsity against
\subsection{Benchmarking hierarchical tree sparsity against
common sparsity penalties}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation.
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with (sparse) group sparsity
imposes a structured $\ell_1/\ell_2$ block norm (with additional $\ell_1$ term)
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
% The stratified and shuffled training data were submitted
% to a nested cross-validation scheme
% for model selection and model assessment.
% In the inner CV layer, the logistic regression estimators
% have been trained in a one-versus-rest design that
% distinguishes each class from
% the respective 17 other classes
% (number of maximal iterations=$100$, tolerance=$0.001$).
% In the outer CV layer, grid search
% selected among candidates for the respective $\lambda$ parameter
% by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
% Importantly, the thus selected sparse logistic regression classifier was
% evaluated on an identical test set in all settings.

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=0.80\textwidth]{../figures/dataratio_trans.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Naive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
(\textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The support recovery of that exemplary task is given
as Pearson correlation (cf. methods).
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance (83.7\% versus 89.0\%).
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{centering}
\end{figure*}

\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).

@@

Three observations have been made.
In the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5,11\% as well as
better support recovery with weight maps already much closer
to the class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
%
Across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representatives among class-responsive predictor groups
due to the $\ell_1$-norm penalization.
%
In the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhance the similarity
to the weight maps even in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator reached this generalization performance
only with bigger input data quantities.
%
\begin{figure*}
\begin{centering}
\includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other.
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}
%
\subsection{Support recovery as a function of region-network emphasis}
% \paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\frac{1}{3}$ to $\eta_g$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

>>>>

\begin{figure*}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.55\textwidth]{../figures/ROC_ALL.png}
\vspace{-0.4cm}
\caption{\textbf{Prediction performance across sparsity priors.}
Comparison of the performance of logistic regression estimators
with 6 different structured and unstructered sparsity penalties
in classifying neural activity from 18 psychological tasks.
The class-wise area under the curve (AUC, \textit{colors})
is obtained on an identical test set.
%
\textit{Hierarchical Tree Sparsity}: Structured
$\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors
exhibited the best out-of-sample performance
($\alpha = 1, \beta = 1$).
\textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
imposed a minimum of relevant brain voxels without
assuming special structure.
\textit{Elastic-Net}: Unstructured logistic regression
with interpolation between $\ell_1$- and $\ell_2$-norm 
imposed an equilibrium between sparsity and model fit.
\textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
norm (with additional $\ell_1$ term)
imposed region compartments, but naive to network structure.
\textit{Trace-norm}: Structured trace-norm penalization
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\end{centering}
\label{fig_sparsities}
\end{figure*}

% \paragraph{Benchmarking hierarchical tree sparsity against
\subsection{Benchmarking hierarchical tree sparsity against
common sparsity penalties}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation
($\alpha = 1, \beta = 1$).
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with (sparse) group sparsity
imposes a structured $\ell_1/\ell_2$ block norm (with additional $\ell_1$ term)
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
% The stratified and shuffled training data were submitted
% to a nested cross-validation scheme
% for model selection and model assessment.
% In the inner CV layer, the logistic regression estimators
% have been trained in a one-versus-rest design that
% distinguishes each class from
% the respective 17 other classes
% (number of maximal iterations=$100$, tolerance=$0.001$).
% In the outer CV layer, grid search
% selected among candidates for the respective $\lambda$ parameter
% by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
% Importantly, the thus selected sparse logistic regression classifier was
% evaluated on an identical test set in all settings.

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=0.80\textwidth]{../figures/dataratio_trans.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Naive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$, \textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The support recovery of that exemplary task is given
as Pearson correlation (cf. methods).
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance (83.7\% versus 89.0\%).
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{centering}
\end{figure*}

\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$)
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).

@@

Three observations have been made.
In the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5,11\% as well as
better support recovery with weight maps already much closer
to the class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
%
Across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representatives among class-responsive predictor groups
due to the $\ell_1$-norm penalization.
%
In the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhance the similarity
to the weight maps even in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator reached this generalization performance
only with bigger input data quantities.
%
\begin{figure*}
\begin{centering}
\includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other
(i.e., $\alpha$ and $\beta$ are changed reciprocally).
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}
%
\subsection{Support recovery as a function of region-network emphasis}
% \paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\beta = \frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\alpha = \frac{1}{3}$ to $\eta_g$ of all network group penalties
(Table \ref{table_reg_net_ratio}).
